RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R

### 2014-03-24 inst/doc/pubs/manuscript.Rmd
updated manuscript
update simmap example in manuscript

[Bollback 2006] provides a widely used stand-alone software implementation
of this method in the software `simmap`, which modified the standard
Newick tree format to express this additional information. This can
break compatibility with other software[^1], and creates a format that
cannot be interpreted without additional information describing this
convention.  By contrast, the NeXML extension is not only backwards
compatible but contains a precise and machine-readable description of
what it is encoding.

@@

At that URL we post a simple description of each term.

@@

Data archiving is increasingly required by journals in evolutionary biology and 
biodiversity [e.g. @Whitlock2011].  The burden of preparing and submitting properly
annotated data to archives continues to be a significant barrier [Tenopir et al. 2011, Stodden 2013],
and many phylogenetic trees upon which studies are based are inaccessible or lost 
to the research community [Drew et al 2013].  `RNeXML` seeks to lower these barriers
by providing functions to immediately archive NeXML files in scientific repositories.

>>>>

[Bollback 2006] provides a widely used stand-alone software implementation
of this method in the software `simmap`, which modified the standard
Newick tree format to express this additional information. This can
break compatibility with other software [^1], and creates a format that
cannot be interpreted without additional information describing this
convention.  By contrast, the NeXML extension is not only backwards
compatible but contains a precise and machine-readable description of
what it is encoding.

@@

At that URL we have posted a simple description of each term.

<!-- How could this be improved? Should we use something more formal? 
We want this example to be something other users could reasonably
do themselves without expertise in RDFa or XML, but also to be a good
model case that doesn't cut corners.  -->

Using this convention we can generate NeXML files containing simmap
data, read those files into R, and convert them back into the `phytools`
package format. These simple functions serve as further illustration of
how `RNeXML` can be used to extend the NeXML standard.  We illustrate
their use briefly here, starting with  loading a nexml object containing
a simmap reconstruction into R:

```{r}
data(simmap_ex)
```

The `get_trees` function can be used to return an `ape::phylo` tree as
usual.  `RNeXML` automatically detects the simmap reconstruction data
and returns includes this in a `maps` element of the `phylo` object,
for use with other `phytools` functions.

```{r}
phy <- nexml_to_simmap(simmap_ex)
```

We can then use various functions from `phytools` designed for simmap
objects, such as the plotting function:

```{r}
library(phytools)
plotSimmap(phy)
```

Likewise, we can convert the object back in the nexml format and write
it out to file to be read by other users.

```{r}
nex <- simmap_to_nexml(phy) 
nexml_write(nex, "simmap.xml")
```

Though other NeXML parsers (for instance, for perl or python) have
not been written explicitly to express simmap data, those parsers will
nonetheless be able to successfully parse this file and expose the simmap
data to the user.

@@

Data archiving is increasingly required by journals in evolutionary biology and 
biodiversity (e.g. [Whitlock2011]).  The burden of preparing and submitting properly
annotated data to archives continues to be a significant barrier ([Tenopir et al. 2011], [Stodden 2010]),
and many phylogenetic trees upon which studies are based are inaccessible or lost 
to the research community [Drew et al 2013].  `RNeXML` seeks to lower these barriers
by providing functions to immediately archive NeXML files in scientific repositories.
